# Model specification

This chapter details the `mrgsolve` model specification format.  

## How / where to write a model
There are two ways to write your model:

### Separate file {#spec-separate-file}
Open a text editor and type the model into a file with name that has the format `<model-name>.cpp`.  This filename format identifies a "name" for your model (`<model-name>`, the "stem" of the file name).  The extension __MUST__ be `.cpp` (`mrgsolve` curently assumes the extension).  Note: this whole file will be read and parsed, so everything in it must be valid `mrgsolve` model specification elements.  

Use the `mread()` function to read and parse this file.  For the model called `mymodel` saved in `mymodel.cpp` (in the current working directory), issue the command:
```{r,eval=FALSE}
mod <- mread("mymodel")
```

`mread()` returns a model object from which you can simulated.



### Inline / "code" {#spec-inline}
Often it is more convenient to write a model right in your `R` script.  The model might look something like this:

```{r}
code <- '
$PARAM CL = 1, VC = 20
$PKMODEL ncmt=1
'
```

Here, we created a character vector of length 1 and saved it to the `R` object called `code`.  The name of this object is irrelevant.  But `code` will be passed into `mrgsolve` as the model definition.  When `mrgsolve` gets a model like this along with a "name" for the model, mrgsolve will write the code to a file called `<model-name>.cpp` and read it right back in as if you had typed the code into this file (section \@ref(spec-separate-file)). 

To parse and load this model, use the `mcode()` command:

```{r,eval=FALSE}
mod <- mcode("mymodel", code)
```

`mcode()` is a convenience wrapper for `mread()`.  `mcode` writes the code to `mymodel.cpp` in `tempdir()`, reads it back in, compiles and loads.

The `mcode` call is equivalent to:
```{r, eval=FALSE}
mod <- mread("mymodel", tempdir(), code)
```

For help, see `?mread` , `?mcode` in the `R` help system after loading `mrgsolve`.

## Variables and Macros
In the following section, we adopt the convention that `CMT` stands for a compartment in the model.

### `ID`
The current subject identifier.

### `TIME`
Gives the time in the curent data set record.  This is usually only used in `$MAIN` or `$TABLE`.  Contrast with `SOLVERTIME`.

### `SOLVERTIME`
Gives the time of the current timestep taken by the solver.  This is can only be used in `$ODE`.  Contrast with `TIME`.

### `EVID`
`EVID` is an event id indicator.  `mrgsolve` recognized the following 
event IDs:

* 0 = an observation record
* 1 = a bolus or infusion dose
* 2 = other type event
* 3 = system reset
* 4 = system reset and dose
* 8 = replace



### `NEWIND`
`NEWIND` is a new individual indicator, taking the following values:

* 0 for the first record of the data set
* 1 for the first record of a given individual
* 2 for all other records

For example:
```{c, eval=FALSE}
$GLOBAL 
int counter = 0;

$MAIN
if(NEWIND <=1) {
  counter  = 0;
}
```


### `ETA(n)`
`ETA(n)` is the value of the subject-level variate drawn from the model `OMEGA` matrix.  `ETA(1)` through `ETA(25)` have default values of zero so they may be used in a model even if appropriate `OMEGA` matrices have not been provided. 

For example:
```{c, eval=FALSE}
$OMEGA
1 2 3

$MAIN
double CL = TVCL*exp(ETA(1));
double VC = TVVC*exp(ETA(2));
double KA = TVKA*exp(ETA(3));
```
Here, we have a 3x3 `OMEGA` matrix. `ETA(1)`, `ETA(2)`, and `ETA(3)` will be populated with variates drawn from this matrix.  `ETA(4)` through `ETA(25)` will be populated with zero. 

### `EPS(n)`
`EPS(n)` holds the current value of the observation-level random variates drawn from `SIGMA`.  The basic setup is the same as detailed in `ETA(n)`.

Example:
```{c,eval=FALSE}
$CMT CENT

$PARAM CL=1, VC=20

$SIGMA
labels=s(ADD,PROP)
25 0.0025

$TABLE
double DV = (CENT/VC)*(1+PROP) + ADD;

```



### `table(name) `
This macro is used in `$TABLE` to insert derived variables or other information into the simulated data set.  

For example:
```{c,eval=FALSE}
$TABLE
table(CP) = CENT/VC;
```
In this example, a column will appear in the simulated data set that will be named `CP` and will take the value of the ratio of the amount in the central compartment divided by the central volume of distribution. 

Notice that the `CP` value only exists in the `table` map.  It may be 
accessed on the right hand side of an expression like this:
```{c,eval=FALSE}
$TABLE
table(CP) = CENT/VC;
table(logCP) = log(table(CP));
```

Note also that the name of the output column will appear exactly 
as it does inside the `table` macro.  So:

```{c,eval=FALSE}
$TABLE
// DON"T DO THIS
// table("CP") = CENT/VC;
// table(ETA(1)) = ETA(1);

// DO THIS
table(CP) = CENT/VC;
table(ETA1) = ETA(1);
```


See: section \@ref(block-table) and also \@ref(block-capture)


### `F_CMT`
For the `CMT` compartment, sets the bioavability fraction for that compartment. 

Example:
```{c,eval=FALSE}
$MAIN
F_CENT = 0.7;
```


### `ALAG_CMT`
For the `CMT` compartment, sets the lag time for doses into that compartment. 

Example:
```{c,eval=FALSE}
$MAIN
ALAG_CENT = 0.25;
```


### `R_CMT`
For the `CMT` compartment, sets the infusion rate for that compartment.  The infusion rate is only set via `R_CMT` when `rate` in the data set or event object is set to `-1`. 

Example:
```{c,eval=FALSE}
$MAIN
R_CENT = 100;
```


### `D_CMT`
For the `CMT` compartment, sets the infusion duration for that compartment.  The infusion duration is only set via `D_CMT` when `rate` in the data set or event object is set to `-2`. 

Example:
```{c,eval=FALSE}
$MAIN
D_CENT = 2;
```


## Code blocks {#code-blocks}

### `$PARAM` \ {#block-param}
Define the parameter list in the current model. 

Example:
```{c, eval=FALSE}
$PARAM CL = 1, VC = 20, KA = 1.2
KM = 25, VMAX = 400, FLAG = 1, WT = 80
SEX = 0, N = sqrt(25)
```

Notes:

* Multiple blocks are allowed
* Values are evaluated by the `R` interperter

See also: section \@ref(block-theta) and \@ref(block-fixed).

### `$FIXED`  \ {#block-fixed}
Like `$PARAM`, `$FIXED` is used to specify `name=value` pairs.  Unlike `$PARAM`, however, the values associated with names in `$FIXED` are not able to be updated.  

By default, names in `$FIXED` are associated with their value through a `C++` preprocessor `#define` statement. 

Usually, `$FIXED` is only used when there are a very large number of parameters ($>$ 100 or 200).  When some of these parameters never need to be updated, you can move them to a `$FIXED` block to get a modest gain in efficiency of the simulation.

Items in `$FIXED` will not be shown when parameters are queried. 

Example:
```{c,eval=FALSE}
$PARAM CL = 2, VC = 20

$FIXED
g = 9.8
```

See also: section \@ref(block-param) and \@ref(block-theta).

Notes:

* Multiple blocks are allowed
* Values are evaluated by the `R` interpreter


### `$CMT`  and  `$INIT`
Declare the names of all compartments in the model. 

* For `$CMT` give the names of compartments; initial values are assumed to be 0
* For `$INIT` give the name and initial value for all compartments

Examples:
```{c,eval=FALSE}
$CMT GUT CENT RESPONSE
```

```{c,eval=FALSE}
$INIT GUT  = 0, CENT = 0, RESPONSE = 25
```

### `$MAIN` {#block-main}
This code block has two main purposes:

* Derive new algebraic relationships between parameters, random, effects and other derived variables
* Set the initial conditions for model compartments

For users who are familiar with NONMEM, `$MAIN` is similar to `$PK`.

`$MAIN` is wrapped into a `C++` function and compiled / loaded by `mrgsolve`.  The `MAIN` function gets called just prior to advancing the system from the current time to the next time for each record in the data set.  `$MAIN` also gets called several times before starting the problem (`NEWIND == 0`) and just prior to simulating each individual (`NEWIND == 1`).  Finally, `$MAIN` gets called every time the model
initial conditions are queried with `init()`.  

New variables may be declared in `$MAIN`.  See section \@ref(new-variables) for details.

Examples:

```{c,eval=FALSE}
$CMT CENT RESP

$PARAM KIN = 100, KOUT = 2, CL = 1, VC = 20

$MAIN

RESP_0 = KIN/KOUT;

double ke = CL/VC;

```


### `$ODE` \ {#block-ode}

Use `$ODE` to define model differential equations.  For all compartments assign the value of the differential equation to `dxdt_CMT` where `CMT` is the name of the compartment.  The `dxdt_` equation may 
be a function of model parameters (via `$PARAM`), the current value of any compartment (`CMT`) or any user-derived variable. 

For example:
```{c,eval=FALSE}
$CMT GUT CENT

$ODE
dxdt_GUT = -KA*GUT;
dxdt_CENT = KA*GUT - KE*CENT;

```

It is important to make sure that there is a `dxdt_` expression defined for every compartment listed in `$CMT` or `$INIT`, even if it is `dxdt_CMT = 0;`

The `$ODE` function is called repeatedly during a simulation run.  So it is wise to do as many calculations as possible outside of `$ODE`, usually in `$MAIN`.  But remember that any calculation that depends on an amount in a compartment and helps determine the `dxdt_` expression in a model must be written in `$ODE`.


New variables may be declared in `$ODE`.  See section \@ref(new-variables) for details.

For example:
```{c,eval=FALSE}
$CMT CENT RESP
$PARAM VC = 100, KE = 0.2, KOUT = 2, KIN = 100
$ODE
double CP = CENT/VC;
double INH = CP/(IMAX+CP)

dxdt_CENT = -KE*CENT;
dxdt_RESP =  KIN*(1 - INH) - RESP*KOUT;

```

If the model needs to refer to the current time, use the `SOLVERTIME` variable.

Notes:

* `$ODE` is written in `C++` syntax; every line must end in `;`
* There may be only one `$ODE` block in a model


### `$TABLE` \ {#block-table}
Use `$TABLE` to interact with parameters, compartment values, and other user-defined variables __after__ the system advances to the next time.  The value associated with any symbol may be inserted into the simulated output with the `table()` macro.  

For example:
```{c,eval=FALSE}
$TABLE
table(CP) = CENT/VC;

```

This construct inserts a column called `CP` into the simulated output, calculated for each record based on the values of the amount in the central compartment (`CENT`) and the parameter (or variable) `VC`.

### `$CAPTURE` \ {#block-capture}
A convenience block to quickly identify variables that should be captured in the simulated output. 

For example:
```{c,eval=FALSE}
$PARAM A = 1, B = 2
$MAIN
double C = 3;
bool yes = true;

$CAPTURE A B C yes
```

This construct will result in 4 additional columns in the simulated output with names `A`, `B`, `C`, and `yes`.  

The `$CAPTURE` block is translated to the following code:
```{c,eval=FALSE}
$PARAM A = 1, B = 2
$MAIN
double C = 3;
bool yes = true;

// $CAPTURE A B C yes 
// inserts the following code at the end of $TABLE
table(A) = A;
table(B) = B;
table(C) = C;
table(yes) = yes;
```

New variables may be declared in `$TABLE`.  See section \@ref(new-variables) for details.


### `$OMEGA` \ {#block-omega}
See `?modMATRIX` for more details about options for this block.

Use this block to enter matrices for subject-level random effects drawn from multivariate normal distribution.  

By default, a __diagonal__ matrix is assumed.  So:

```{c, eval=FALSE}
$OMEGA
1 2 3
```
will generate a 3x3 omega matrix.

A __block__ matrix may be entered by using `block=TRUE`. So:
```{c, eval=FALSE}
$OMEGA block = TRUE
0.1 0.02 0.3
```
will generate a 2x2 matrix with covariance 0.02.

A 2x2 matrix where the off-diagonal element is a correlation, not a covariance can be specified like this:
```{c,eval=FALSE}
$OMEGA correlation = TRUE
0.1 0.67 0.3
```

Here, the correlation is 0.67. `mrgsolve` will calculate the covariances and substitute these values.  The matrix will be stored and used with these covariances, not the correlation.

A name can be assigned to each matrix:
```{c, eval=FALSE}
$OMEGA name = "PK", block=TRUE
0.2 0.02 0.3

$OMEGA name="PD"
0.1 0.2 0.3 0.5
```
to distinguish between multiple `$OMEGA` blocks and to facilitate updating later. The model in the preceeding example will have two 
`$OMEGA` matrices: 2x2 and 4x4.


Notes:

* Multiple `$OMEGA` blocks are allowed


### `$SIGMA` \ {#block-sigma}
See `?modMATRIX` for more details about options for this block.

Use this block to enter matrices for within-subject random effects drawn from multivariate normal distribution.  

The `$SIGMA` block functions like the `$OMEGA` block.  See `$OMEGA` for details. 

### `$SET`
Use this code block to set different options for the simulation. Use a `name=value` format, where `value` is evaluated by the `R` interpreter.  

Most of the options that can be entered in `$SET` are passed to `update`. 

For example:
```{c,eval=FALSE}
$SET end = 240, delta=0.5, req=s(RESP)
```

Here, we set the simulation `end` time to 240, set the time difference between two
adjacent time points to 0.25 time units, and request only the `RESP`onse compartment in the simulated output.


### `$GLOBAL`
The `$GLOBAL` block is for writing `C++` code that is outside of `$MAIN`, `$ODE`, and `$TABLE`.

There are no artificial limit on what sort of `C++` code can go in `$GLOBAL`.  However there are two more-common uses:

1. Write `#define` preprocessor statements
1. Define global variables, usually variables other than `double`, `bool`, `int` (see \@ref(new-variables))


__Preprocessor directives__
Preprocessor  `#define` directives are direct substibutions that the `C++` preprocessor makes prior to compiling your code.  

For example:
```{c,eval=FALSE}
$GLOBAL
#define CP (CENT/VC)
```
When this preprocessor directive is included, everywhere the preprocessor finds a `CP` token it will substitute `(CENT/VC)`.  Both `CENT` and `VC` must be defined and the ratio of `CENT` to `VC` will be calculated depending on whatever the current values are.  Notice that we included parentheses around `(CENT/VC)`.  This makes sure the ratio between the two is taken first, before any other operations involving `CP`.  

__Declaring global variables__
Sometimes, you may wish to use global variables and have more control over how they get declared.  

```{c,eval=FALSE}
$GLOBAL
bool cure = false;
```
With this construct, the boolean variable `cure` is declared and defined right as the model is compiled.  


### `$PKMODEL`
This code block implements a one- or two-compartment PK model where the system is calculated by algebraic equations, not ODEs.  `mrgsolve` handles the calculations and an error is generated if both `$PKMODEL` and `$ODE` blocks are included in the same model specification file.

This is an options-only block. The user must specify the number of compartments (1 or 2) to use in the model as well as whether or not to include a depot dosing compartment.  See `?PKMODEL` for more details about this block, including specific requirements for symbols that must be defined in the model specification file.

The `$CMT` or `$INIT` block must also be included with an appropriate number of compartments.  Compartment names, however, may be determined by the user.

Example:
```{c, eval=FALSE}
$CMT GUT CENT PERIPH
$PKMODEL ncmt=2, depot=TRUE
```


### `$THETA`  \ {#block-theta}
Use this code block as an efficient way to add to the parameter list where 
names are determined by a prefix and a number.  By default, the prefix is `THETA` and the number sequentially numbers the input values. 

For example:

```{c,eval=FALSE}
$THETA
0.1 0.2 0.3
```
is equivalent to
```{c,eval=FALSE}
$PARAM THETA1 = 0.1, THETA2 = 0.2, THETA3 = 0.3
```

See also: \@ref(block-param) .


### `$NMXML`  \ {#block-nmxml}
The `$NMXML` block lets you read and incorporate results from a NONMEM run into 
your `mrgsolve` model.  From the NONMEM run, `THETA` will be imported into your parameter
list (see \@ref(block-param) and \@ref(component-param)), `OMEGA` will be captured as an `$OMEGA` block (\@ref(block-omega)) and `SIGMA` will be captured as a `$SIGMA` block (\@ref(block-sigma)).  Users may optionally omit any one of these from being imported.


__An example__

There is a NONMEM run embedded in the `mrgsolve` package 
```{r}
path <- file.path(path.package("mrgsolve"),"nonmem")
list.files(path, recursive=TRUE)

```

 
We can create a `mrgsolve` control stream that will import `THETA`, `OMEGA` and 
`SIGMA` from that run using the `$NMXML` code block.

```{r}
code <- '
$NMXML
run = 1005
project=path
theta=TRUE, omega=TRUE,sigma=TRUE
olabels = s(ECL, EVC, EKA)
slabels = s(PROP, ADD)

$MAIN
double CL = THETA1*exp(ECL);
double V2 = THETA2*exp(EVC);
double KA = THETA3*exp(EKA);
double Q = THETA4;
double V3 = THETA5;

$PKMODEL ncmt=2, depot=TRUE

$CMT GUT CENT PERIPH

$TABLE
table(CP) = (CENT/V2)*(1+PROP) + ADD/5;

$SET delta=4, end=96
'
```




__NOTE__: in order to use this code, we need to install the `XML` package.

```{r}
mod <- mcode("nmxml", code, quiet=TRUE)

mod

param(mod)

revar(mod)

```
An infusion into `GUT`:
```{r}
set.seed(2922)
#+
mod %>%
  ev(amt=1000, rate=1000/36, cmt="GUT") %>%
  mrgsim(nid=10) %>%
  plot
```




## Derive new variables {#new-variables}
New `C++` variables may be derived in `$GLOBAL`, `$MAIN`, `$ODE` and `$TABLE`.  Because these are `C++` variables, the type of variable being used must be declared.  For the vast majority of applications, the `double` type is used (double-precision numeric value. 

```{c,eval=FALSE}
$MAIN
double CLi = TVCL*exp(ETA(1));
```

We want `CLi` to be a numeric value, so we use `double`.  To derived a `boolean` variable, write
```{c,eval=FALSE}
$MAIN
bool cure = false;
```

When variables of the type `double`, `int`, and `bool` are declared and initialized in `$MAIN`, `$ODE`, `$TABLE`, `mrgsolve` will detect those declarations, and modify the code so that the variables are actually declared in `$GLOBAL` not in `$MAIN`, `$ODE`, or `$TABLE`.  This is done so that variables declared in one code block can be read and modified in another code block.  

For example, the following code:
```{c,eval=FALSE}
$MAIN
double CLi = TVCL*exp(ETA(1));
```

gets translated to:
```{c,eval=FALSE}
$GLOBAL
double CLi;

$MAIN
CLi = TVCL*exp(ETA(1));
```

This way, we can still read `CLi` in `$TABLE`:
```{c,eval=FALSE}
$MAIN
double CLi = TVCL*exp(ETA(1));
double VCi = TVVC*exp(ETA(2));

$TABLE
table(KEi) = CLi/VCi;
```

To declare a variable that is local to a particular code block:

```{c,eval=FALSE}
$MAIN
localdouble CLi = TVCL*exp(ETA(1));

```










